---
title: "Hmwk 3"
author: "Hunt Jones"
date: "2/4/2022"
output: html_document
---

## Univariate Assignment

Read in tree data

```{r}

library(ggplot2)
library(gridExtra)
library(scatterplot3d)
library(MASS)
library(car)

```

```{r}
# read in directly from website: 
trees <- read.csv('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/treedata_subset.csv')

str(trees)
```

Examine this dataset and see how the data is structured, see function 'str' 

The contents of the metadata file (./data/tree_metadata.txt(https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/tree_metadata.txt)) is provided here:


The dataset includes tree abundances from a subset of a vegetation database of Great Smoky Mountains National Park (TN, NC).

* plotID: unique code for each spatial unit (note some sampled more than once)
* date: when species occurrence recorded
* plotsize: size of quadrat in m2
* spcode: unique 7-letter code for each species
* species: species name
* cover: local abundance measured as estimated horizontal cover (ie, relative area of shadow if sun is directly above) classes 1-10 are: 1=trace, 2=0-1%, 3=1-2%, 4=2-5%, 5=5-10%, 6=10-25%, 7=25-50%, 8=50-75%, 9=75-95%, 10=95-100%
* utme: plot UTM Easting, zone 17 (NAD27 Datum)
* utmn: plot UTM Northing, zone 17 (NAD27 Datum)
* elev: elevation in meters from a digital elevation model (10 m res)
* tci: topographic convergence index, or site "water potential"; measured as the upslope contributing area divided by the tangent of the slope angle (Beven and Kirkby 1979)
* streamdist: distance of plot from the nearest permanent stream (meters)
* disturb: plot disturbance history (from a Park report); CORPLOG=corporate logging; SETTLE=concentrated settlement, VIRGIN="high in virgin attributes", LT-SEL=light or selective logging
* beers: transformed slope aspect ('heat load index'); 0 is SW (hottest), 2 is NE (coolest)



Above shows a map of the regional and local location of the elevational transects included in the dataset (from Fridley 2009 (http://plantecology.syr.edu/fridley/Fridley2009_jamc.pdf)).


1. Carry out an exploratory analysis using the tree dataset. CHECK

Metadata for the tree study can be found here(../data/tree_metadata.txt). 

Specifically, I would like you to develop and compare models for species cover for a habitat
generalist Acer rubrum


and a habitat specialist Abies fraseri 



Because this dataset includes both continuous and discrete explanatory variables
use the function Anova in the packages car as such


```{r}
explore <- trees[c( which(trees$species == "Abies fraseri"), which(trees$species == "Acer rubrum")), ]
abies <- trees[which(trees$species == "Abies fraseri"), ]
acer <- trees[which(trees$species == "Acer rubrum"), ]
```

```{r}
abies$disturb <- as.factor(abies$disturb) # treat disturbance as a factor

abies$disturb <- factor(abies$disturb, levels = c("VIRGIN", "CORPLOG", "LT-SEL", "SETTLE"))
# this reorders the disturbance factors so that when we run the 'summary' function,
# CORPLOG, LT-SEL, and SETTLE are each compared to VIRGIN (VIRGIN essentially
# becomes the null hypothesis.

plot(cover ~ elev, data = abies)
plot(cover ~ tci, data = abies)
plot(cover ~ streamdist, data = abies)
plot(cover ~ beers, data = abies)        # plot each variable independently to 
plot(cover ~ disturb, data = abies)      # look for signal in graphs

# based on the output of these graphs, prior to making my model, I would expect
# to see a significant effect of elevation on cover, and another one by distance
# to the nearest stream, though I would expect the effect of elevation on cover 
# to be stronger. I expect no significant effect of beers or tci on cover.
# For the disturbance levels, expect that there might be a significant difference
# between the VIRGIN and CORPLOG data, but there is such high variance in the 
# VIRGIN data that the difference may not be statistically significant.

```

```{r}
abies_mod <- lm(cover ~ elev + tci + streamdist + beers + disturb, data = abies)
abies_mod   # creates our univariate model with cover as the response, which 
            # evaluates the effect of each independent variable individually 
            # and compares it to the others
```

```{r}
Anova(abies_mod, type = 3)
```

```{r}
summary(abies_mod)
```

```{r}
acer$disturb <- as.factor(acer$disturb)

acer$disturb <- factor(acer$disturb, levels = c("VIRGIN", "CORPLOG", "LT-SEL", "SETTLE"))

plot(cover ~ elev, data = acer)
plot(cover ~ tci, data = acer)
plot(cover ~ streamdist, data = acer)
plot(cover ~ beers, data = acer)
plot(cover ~ disturb, data = acer)

# my expectations for the acer data are quite different than that of Abies fraseri.
# I expect no significant effect of any of the independent variables included
# on cover, and I expect to find no significant difference in cover among the
# samples taken from areas with different disturbance levels.

```

```{r}
acer_mod <- lm(cover ~ elev + tci + streamdist + beers + disturb, data = acer)
acer_mod 
```

```{r}
Anova(acer_mod, type=3)

```

```{r}
summary(acer_mod)

```


This will estimate partial effect sizes, variance explained, and p-values for 
each explanatory variable included in the model. 

Compare the p-values you observe using the function 'Anova' to those generated
using 'summary'. 
All p-values are the same for both Anova and summary except for that of the 
disturbance factor. Anova runs one test comparing all disturbance types at once,
while summary compares each disturbance type to the VIRGIN group, which serves
as the null hypothesis.

For each species address the following additional questions:

* how well does the exploratory model appear to explain cover?
    Abies fraseri: the exploratory model indicates that elevation is the best
    variable to explain cover. Elevation was the only explanatory variable with 
    a significant relationship to cover. The main reason that this model shows
    promise, however, is that the r-squared is fairly high, especially for data
    collected in the field rather than a controlled lab setting. However, the
    adjusted r-squred value is lower than the r-squared value by about 0.08, 
    which means that some overfitting has occurred and this model would probably
    be a better model if it included fewer explanatory variables.
    
    Acer rubrum: the exploratory model for acer rubrum does not explain cover
    very well. While several explanatory variables were found to have a 
    significant relationship with cover, the r-squared value was very low,
    meaning that the explanatory variables do not explain much of the variance
    in the cover data. On top of that, the adjusted r-squared is lower than the
    r-squared. Although it is only slightly lower, the r-squared value here is
    so low that even a small decrease might indicate overfitting.
    
* which explanatory variables are the most important?
    the most important explanatory variable in each model appears to be elevation.
    
* do model diagnostics indicate any problems with violations of OLS assumptions?
    
```{r}
par(mfrow = c(2,2))
plot(abies_mod)
# The data appear to be linear, the residuals are normally distributed. However,
# homoscedasticity is questionable because the data re not evenly spread. The 
# data do not appear to be leveraged, as the standardized residuals are within
# 3 standard deviations, meaning that no outliers exist that might alter the 
# interpretation of our data, though if I am being honest, I still have a hard 
# time understanding how to assess this violation.

```

```{r}
par(mfrow = c(2,2))
plot(acer_mod)

# The residuals vs fitted appear to indicate that the data is linear, though
# they are not perfect. The residuals appear to be normally distributed. The 
# data to not appear to have homoscedasticity. And the data appear to be 
# levearaged with several outliers.

```
    
* are you able to explain variance in one species better than another, 
  why might this be the case?
  
  Yes, we can explain variance in Abies fraseri better. This might be because
  fewer of this species of tree were observed. Often, when there is a large
  sample size, significant relationships might be identified despite low 
  r-squared.




2. You may have noticed that the variable cover is defined as 
positive integers between 1 and 10. and is therefore better treated
as a discrete rather than continuous variable. 
Re-examine your solutions to the question above but from the
perspective of a General Linear Model (GLM) with a Poisson error term
(rather than a Gaussian one as in OLS). 
The Poisson distribution generates integers 0 to positive infinity so this may provide a good first approximation. 
Your new model calls will look as follows:

```{r}
acer_poi <- glm(cover ~ elev + tci + streamdist + beers + disturb, data = acer, family = poisson)
acer_poi

Anova(acer_poi, type=3)

summary(acer_poi)
```

```{r}
abies_poi <- glm(cover ~ elev + tci + streamdist + beers + disturb, data = abies, family = poisson)
abies_poi

Anova(abies_poi, type=3)

summary(abies_poi)

```

For assessing the degree of variation explained you can use a 
pseudo-R-squared statistic (note this is just one of many possible)

```{r}
pseudo_r2 = function(glm_mod) {
                1 -  glm_mod$deviance / glm_mod$null.deviance
}

pseudo_r2(acer_poi)

pseudo_r2(abies_poi)
```

Compare your qualatitive assessment of which variables were most important in each model. 
Does it appear that changing the error distribution changed the results much? In what ways?

Elevation was the most important variable in every model. Changing the error 
distribution did change the results slightly, but not in a meaningful way, in 
my opinion. Elevation was still the only variable with a significant relationship
to cover in the Abies fraseri model, and the variable of significance in the 
Acer rubrum model were the same in the lm and glm models, though the p-values
were slightly different. Using the pseudo r-squared value as justification, it
appears that the glm models were better, but only by a little bit.

3. Provide a plain English summary (i.e., no statistics) of what you have
found and what conclusions we can take away from your analysis?

What we have found is that elevation appears to be the primary factor of value
in predicting any effect on cover. For Abies fraseri, using elevation as a 
predictor of cover will produce fairly reliable results. However, using elevation
as a predictor of cover for Acer rubrum will not provide very reliable results.

4. (optional) Examine the behavior of the function 'stepAIC()' using the 
exploratory models developed above. This is a very simple and not very
robust machine learning stepwise algorithm that uses AIC to select a 
best model. By default it does a backward selection routine. 

I am still a bit shaky on interpreting AIC results. But, the basics of what I 
know are that lower AIC's indicate better models. Each of the results here 
appears to confirm what I found in my lm and glm models earlier, in that elevation
appears to be the most important variable in the model. The best model for 
Acer rubrum appears to be a glm model using elevation and tci as the explanatory
variables, and the best model for Abies fraseri appears to be the same (a glm
model with tci and elevation as the explanatory variables).

```{r}
stepAIC(acer_mod)
stepAIC(acer_poi)
stepAIC(abies_mod)
stepAIC(abies_poi)

```



5. (optional) Develop a model for the number of species in each site 
(i.e., unique plotID). This variable will also be discrete so the Poisson
may be a good starting approximation. Side note: the Poisson
distribution converges asymptotically on the Gaussian distribution as the 
mean of the distribution increases. Thus Poisson regression does not differ
much from traditional OLS when means are large. 