---
title: "HW 4 - Multivariate"
author: "Hunt Jones"
date: "2/10/2022"
output: html_document
---
---
layout: page
title: multivariate models
---

For this assignment will be analyzing data on the Vegetation
and Environment in Dutch Dune Meadows.

To import the data and read the metadata run the following:

```{r}
library(vegan)
data(dune)
data(dune.env)
?dune
?dune.env
```

1. Conduct an indirect ordination on the dune plant community. Specifically,
visually examine a NMDS plot using the bray-curtis distance metric. Below is 
some code to help you develop a potential plot that emphasizes the role of the
environmental variable "Moisture". Describe how you interpret the
graphic. What is the goal of creating such a plot? Does this analysis suggest
any interesting findings with respect to the dune vegetation?

    ```{r}
    dune_mds <- metaMDS(dune, distance = "bray")
    
    dune.env$Moisture <- as.numeric(dune.env$Moisture)
    
    
    plot(dune_mds, type='n')
    text(dune_mds, 'sp', cex=.5)
    # generate vector of colors 
    color_vect = rev(terrain.colors(6))[-1]
    points(dune_mds, 'sites', pch=19, 
           col=color_vect[dune.env$Moisture])
    legend('topright', paste("Moisture =", 1:5, sep=''), 
           col=color_vect, pch=19)
    
    # Based on this graph, it appears that sites that are more similar to one 
    # another tend to have the same moisture, where dryer sites are similar to 
    # other dry sites and wetter sites are similar to other wet sites. What this
    # means is that the plant community is largely determined by moisture.
    ```
    
    
```{r}

stressplot(dune_mds, main = "Shepherd Plot")

gof=goodness(dune_mds)
plot(dune_mds, type="t", main="Goodness of Fit")
points(dune_mds, display="sites", cex=gof*200, col=color_vect[dune.env$Moisture])
```

2. Carry out a direct ordination using CCA in order to test any potential 
hypotheses that you developed after examining the MDS plot. Specifically,
carry out a test of the entire model (i.e., including all constrained axes)
and also carry out tests at the scale of individual explanatory variables
you included in your model if you included more than one variable. Plot your 
results.

```{r}
library(vegan)
dune_cca <- cca(dune)
dune_cca
summary(dune_cca)
summary(dune_cca, scaling=1)   # CCA using entire model.
```

```{r}
evplot <- function(x) {
  par(mfrow=c(2,1))
  barplot(x, main="Eigenvalues", col="bisque", las=2)
  abline(h=mean(x), col="red")
  legend("topright", "Average Eigenvalue", lwd=1, col=2, bty="n")
}
brkstk <- function(x) {
  n <- length(x)
  bsm <- data.frame(j=seq(1:n), p=0)
  bsm$p[1] <- 2/n
  for(i in 2:n) {
    bsm$p[i]=bsm$p[i-1] + (1/(n+1-i))
  }
  bsm$p <- 200*bsm$p/n
  bsm
  par(mfrow=c(2,1))
  barplot(t(cbind(200*x/sum(x), bsm$p[n:1])), beside=TRUE, main="% variance", col=c("bisque",2), las=2)
  legend("topright", c("% eigenvalue", "broken stick model"), pch=15, col=c("bisque", 2), bty="n")
}

```

```{r}
(ev2 <- dune_cca$CA$eig)
evplot(ev2)
brkstk(ev2)
```

```{r}
par(mfrow=c(1,2))
plot(dune_cca, scaling=2, main="CA plant sites - biplot scaling two")
dune_cca_env <- envfit(dune_cca, dune.env)
plot(dune_cca_env) # Generates plot indicating importance of each variable.
                   # It is pretty clear that Moisture is the most important
                   # by looking at this plot, however there is a LOT of noise.
```

```{r}
dune_cca2 <- cca(dune ~ ., dune.env)

plot(dune_cca2)

cca_dune_stepfwd <- ordistep(cca(dune~1, data=dune.env), scope=formula(dune_cca2), directions="forward", pstep=1000) # This cca model indicates, similar to the above, that moisture
            # is the variable that best explains the similarity of sites to one
            # another. However, this model goes on to suggest that Management is
            # also important; a finding that was not obvious (at least to myself)
            # thus far. Now to build a parsimonious model, we will incorporate
            # only Moisture and Management.

```

```{r}
(pars_dune_cca <- cca(dune ~ Moisture + Management, data=dune.env))
# Interestingly, this parsimonious model includes only 4 constrained axes, which
# is the number of CCA axes that was suggested to be included by the Average
# Eigenvalue and broken stick model graphs above.
anova.cca(pars_dune_cca, step=1000)
anova.cca(pars_dune_cca, step=1000, by="axis")

vif.cca(dune_cca2)
vif.cca(pars_dune_cca)
```


3. Do your two analyses agree with one another or complement one another or do
these two analyses seem to be suggesting different take home messages? Which
analysis do you find to be more useful?

These analyses appear to complement one another. Both suggest that Moisture has
the largest effect on site similarity out of all variables available. I suppose
NMDS seemed more straightforward/easy to read. However, CCA, once interpreted
correctly yields more information.
