---
title: "R_HW_2"
author: "Hunt Jones"
date: "1/18/2022"
output: html_document
---
R intermediate
Dan McGlinn

January 15, 2016

Premature optimization is the root of all evil -- Donald Knuth

The humble for loop is often considered distasteful by seasoned programmers because it is inefficient; however, the for loop is one of the most useful and generalizable programming structures in R. If you can learn how to construct and understand for loops then you can code almost any iterative task. Once your loop works you can always work to optimize your code and increase its efficiency.

Before attempting these exercises you should review the lesson R intermediate in which loops were covered.

Examine the following for loop, and then complete the exercises

```{r}
data(iris)
head(iris)
```
##   Sepal.Length Sepal.Width Petal.Length Petal.Width Species
## 1          5.1         3.5          1.4         0.2  setosa
## 2          4.9         3.0          1.4         0.2  setosa
## 3          4.7         3.2          1.3         0.2  setosa
## 4          4.6         3.1          1.5         0.2  setosa
## 5          5.0         3.6          1.4         0.2  setosa
## 6          5.4         3.9          1.7         0.4  setosa

```{r}
sp_ids = unique(iris$Species)

output = matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1) 
#output will create a matrix where the rows are unique species and the columns
#will include all variables from iris, except for the "Species" column.

rownames(output) <- sp_ids
#rownames(output) tells R to name the rows with the unique species names in
#iris.
colnames(output) <- names(iris[ , -ncol(iris)])
#colnames(output) is a command that allows us to tell R to name the columns in 
#our new data.frame using the names of the columns in "iris," excluding the 
#column for "species" at the end.


for(i in seq_along(sp_ids)) {  
  #set up for loop where i represents the length of the number of unique species
    iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species) 
    #FOR each unique species, create a data.frame that subsets all the data for 
    #that species from the original iris data.frame.
    
    for(j in 1:(ncol(iris_sp))) {
      #set up for loop where j represents each column in each of the
      #iris_sp subsets.
        x = 0
        y = 0
        #begin at zero (reset each time a new species is reached at end of 
        #subset)
        if (nrow(iris_sp) > 0) {
          #IF loop; if the row on which data is found is greater then zero, then
          #do the following...
            for(k in 1:nrow(iris_sp)) {
              #k represents each row in each iris_sp subset
                x = x + iris_sp[k, j]
                #x = sum of all rows along each column
                y = y + 1
                #y = total number of rows
            }
            output[i, j] = x / y 
            #output[i,j] = x/y tells R that "output" should return the value of 
            #the sum of each column divided by the number of rows, for each 
            #species.
        }
    }
}
output
```
##            Sepal.Length Sepal.Width Petal.Length Petal.Width
## setosa            5.006       3.428        1.462       0.246
## versicolor        5.936       2.770        4.260       1.326
## virginica         6.588       2.974        5.552       2.026
Excercises
Iris loops
Describe the values stored in the object output. In other words what did the loops create?

The loops created a new data.frame that shows each unique species as a row, and the mean of each measurement for that species in the four columns following, having removed the column for species which would now be redundant.

The loops calculated the mean of each variable (sepal length, sepal width, petal length, and petal width) for each of the 3 unique species in the dataset. 

Describe using pseudo-code how output was calculated, for example,

Loop from 1 to length of species identities
   Take a subset of iris data 
   Loop from 1 to number of columns of the iris data
      If ... occurs then do ...  
The variables in the loop were named so as to be vague. How can the objects output, x, and y be renamed such that it is clearer what is occurring in the loop.

  Output could be more descriptive by being written as something like 
  Trait_means. X could be more descriptive if it was written as trait_sum. Y 
  could be written as n or # of data.

It is possible to accomplish the same task using fewer lines of code? Please suggest one other way to calculate output that decreases the number of loops by 1.
  
  Yes, see code below. I decreased the number of loops by 2, which is not exactly
  what the directions said to do, but it is technically more efficient I think.
  
```{r}
idx = unique(iris$Species)

trait_mn = matrix(0, nrow=length(idx), ncol=ncol(iris)-1) 

rownames(trait_mn) <- idx

colnames(trait_mn) <- names(iris[ , -ncol(iris)])

for(i in seq_along(idx)) {  
    spec = subset(iris, subset=Species == idx[i], select=-Species) 
    for(j in 1:(ncol(spec))) {
      
       trait_mn[i,j] <- mean(spec[i,j])
       
        }
    }
trait_mn
      

```

Sum of a sequence
You have a vector x with the numbers 1:10. Write a for loop that will produce a vector y that contains the sum of x up to that index of x. So for example the elements of x are 1, 2, 3, and so on and the elements of y would be 1, 3, 6, and so on.

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
y <- 0
a_vector_y <- 0
for(i in x) {
  y <- y+i
   a_vector_y[i] <- y
}
a_vector_y
```

Modify your for loop so that if the sum is greater than 10 the value of y is set to NA
```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
y <- 0
a_vector_y <- 0
for(i in x) {
  y <- y+i
   a_vector_y[i] <- y
  
  if(a_vector_y[i] < 10){
    print(y)
  }
    else {
      print(NA)
    }
 
}
```

Place your for loop into a function that accepts as its argument any vector of arbitrary length and it will return y.

```{r}
vect_y <- function(a) {
  y <- 0
  for(i in a) {
    y <- y+i
    
    
  }
}
  
ronda <- c(1:5)
micky <- c(1:7)
vect_y(micky)

as <- vect_y(ronda)

```

(Optional)Fibonacci numbers and Golden ratio
Fibonacci numbers are a sequence in which a given number is the sum of the precedding two numbers. So starting at 0 and 1 the sequence would be
0, 1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, ...

Write and apply a simple R function that can accomplish this task with a for loop. Then write a function that computes the ratio of each sequential pair of Fibonacci numbers. Do they asympoticly approch the golden ratio (1 + sqrt(5)) / 2) ?
