---
title: "Spatial Models HW"
author: "Hunt Jones"
date: "2/22/2022"
output: html_document
---
---
layout: page
---

## Spatial Modeling Assignment

```{r}
library(vegan)


data(BCI)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
```

1) Examine if there is evidence of spatial dependence in a rare and a common
species in the BCI tree dataset

  Based on the graphs of geographic distance and distance in abundance of 
  individuals of Cassipourea guianensis and Symphonia globulifera below, it 
  appears that rare species are spatially independent and common species are 
  spatially dependent. However, only one species was tested each time. When I 
  reran my code using a different common species and a 
  different rare species, the results suggested that both were spatially 
  independent, and when I ran the code again using a third rare species, the 
  results suggested that its distribution was not spatially 
  independent. Therefore, the evidence suggests that both rare and common 
  species can have spatial dependence, but that spatial dependence can be 
  determined on a species by species basis, and cannot be generalized to all 
  rare or all common species.
  Mantel test and corellogram support my conclusion based on the graphs of 
  distance in abundance and population size for Cassipourea guianensi and 
  Symphonia globulifera. C. guianensis does not have spatial dependence, while
  S. globulifera does.

```{r}
plot(BCI_xy)

sp_abund <- colSums(BCI)
hist(sp_abund)
hist(log10(sp_abund)) # log transformation for normal distribution
quantile(sp_abund)
quantile(log10(sp_abund))
which(sp_abund >25 & sp_abund < 27)
which(sp_abund > 10^1.93 & sp_abund < 10^1.95)

rare <- BCI[ , 190]
common <- BCI[ , 32]

plot(BCI_xy, cex = rare/max(rare))
plot(BCI_xy, cex = common/max(common))

geod <- dist(BCI_xy)
rared <- dist(rare)
commd <- dist(common)

maxdist <- max(geod)/2

par(mfrow=c(1,2))
plot(geod, rared, main = 'rare spp')
lines(lowess(geod, rared), lw=2, col='red')
plot(geod, commd, main = 'common spp')
lines(lowess(geod, commd), lw=2, col='red')


# run mantel test and corellogram 

rare_mantel <- mantel(rared, geod)
rare_mantel

common_mantel <- mantel(commd, geod)
common_mantel

rare_corlgrm <- mantel.correlog(rared, geod)

common_corlgrm <- mantel.correlog(commd, geod)


par(mfrow=c(1,2))
plot(rare_corlgrm)
mtext(side=3, 'Rare Species')
abline(v = maxdist, col='red', lwd=3, lty=2)
plot(common_corlgrm)
mtext(side=3, 'Common Species')
abline(v = maxdist, col='red', lwd=3, lty=2)


```


2) Build two generalized linear models to predict the abundance of the species
*Drypetes standleyi* using the abundance of other tree species in the study site.
Specifically examine the following species as predictor variables:


```{r}
sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")

```

```{r}
library(nlme)

pred_sp <- BCI[ , sp_ids]

resp_sp <- BCI[ , 'Drypetes.standleyi']

sp_dat <- data.frame(pred_sp, resp_sp, BCI_xy)

pred_sp_log <- log10(pred_sp+1)

resp_sp_log <- log10(resp_sp+1)

sp_dat_log <- data.frame(pred_sp_log, resp_sp_log, BCI_xy)

sp_mod1 <- gls(resp_sp ~ Xylopia.macrantha, data=sp_dat)
summary(sp_mod1)

sp_mod2 <- gls(resp_sp ~ . -(x+y), data=sp_dat)
summary(sp_mod2)

sp_mod2_log <- gls(resp_sp_log ~ . -(x+y), data=sp_dat_log)
summary(sp_mod2)

```
Note renaming the species ids to something a little easier to work with like
"sp_a", "sp_b" will make model construction a little less cumbersome

* Model 1: only include a single species as a predictor variable
  

* Model 2: include all of the species as predictor variables

  Model 1, with only one species as a predictor, shows strong spatial 
  correlation. The semivariogram first reaches 1.0 at a distance somewhere 
  between 450 to 500, and there appears to be a nugget at about 0.3. However, 
  Model 2, with all species included as predictors, does not appear to show 
  spatial correlation. Using a different spatial correlation structure might yield
  different results for model 2. Of note, the semivariogram in model 2 does not
  appear to fit the data as well as in model 1. While the semivariogram in model
  2 did not reach 1, there are data at or slightly above 1, beginning at distance
  of about 410, and based on visual inspection, there is a nugget at about 0.7.
  
```{r}
plot(Variogram(sp_mod1, form = ~x+y))

plot(Variogram(sp_mod2, form = ~x+y))

plot(Variogram(sp_mod2_log, form = ~x+y))

```

With both models examine the spatial dependence of the residuals using the 
function `Variogram`. Model the spatial dependence in the residuals using one
of the error structures available. 
  

```{r}
sp_mod1_spher <- update(sp_mod1,
corr=corSpher(c(475,0.3), form = ~x+y, nugget=T))
summary(sp_mod1_spher)

sp_mod1_exp <- update(sp_mod1,
corr=corExp(form = ~ x+y))
summary(sp_mod1_exp)

sp_mod1_exp_nug <- update(sp_mod1, corr=corExp(form = ~ x+y, nugget=0.3))
summary(sp_mod1_exp_nug)

plot(Variogram(sp_mod1_spher, resType = "n"))

plot(Variogram(sp_mod1_exp, resType = "n"))

anova(sp_mod1, sp_mod1_spher, sp_mod1_exp, sp_mod1_exp_nug)

```


```{r}
sp_mod2 <- gls(resp_sp ~ Cordia.lasiocalyx+Hirtella.triandra+Picramnia.latifolia+Quassia.amara+Tabernaemontana.arborea+Trattinnickia.aspera+Xylopia.macrantha, data = sp_dat)
summary(sp_mod2)

sp_mod2_spher <- update(sp_mod2, corr=corSpher(c(410, 0.7), form =~x+y, nugget=T))
summary(sp_mod2_spher)

sp_mod2_exp <- update(sp_mod2, corr=corExp(form =~x+y, nugget=T))
summary(sp_mod2_exp)

plot(Variogram(sp_mod2_spher, form = ~x+y))

plot(Variogram(sp_mod2_exp, form = ~x+y, maxDist=maxdist))

plot(sp_mod2,resid( ., type="n")~fitted(.),abline=0)

qqnorm(sp_mod2,~resid(.,type="n"))
```

```{r}
sp_mod2_log <- gls(resp_sp_log ~ Cordia.lasiocalyx+Hirtella.triandra+Picramnia.latifolia+Quassia.amara+Tabernaemontana.arborea+Trattinnickia.aspera+Xylopia.macrantha, data = sp_dat_log)
summary(sp_mod2_log)

sp_mod2_spher_log <- update(sp_mod2_log, corr=corSpher(c(410, 0.7), form =~x+y, nugget=T))
summary(sp_mod2_spher_log)

sp_mod2_exp_log <- update(sp_mod2_log, corr=corExp(form =~x+y))
summary(sp_mod2_exp_log)

sp_mod2_rat_log <- update(sp_mod2_log, corr=corRatio(form = ~x+y))
summary(sp_mod2_rat_log)

sp_mod2_rat_log_nug <- update(sp_mod2_log, corr=corRatio(form = ~x+y, nugget=T))
summary(sp_mod2_rat_log_nug)

plot(Variogram(sp_mod2_spher_log, form = ~x+y))

plot(Variogram(sp_mod2_exp_log, form = ~x+y, maxDist=maxdist))

plot(Variogram(sp_mod2_rat_log, maxDist = maxdist))

plot(Variogram(sp_mod2_rat_log_nug, maxDist = maxdist))

plot(Variogram(sp_mod2_rat_log_nug, resType="n", maxDist = maxdist))

plot(sp_mod2_log,resid( ., type="n")~fitted(.),abline=0)

qqnorm(sp_mod2_log,~resid(.,type="n"))

anova(sp_mod2_log, sp_mod2_exp_log, sp_mod2_spher_log, sp_mod2_rat_log, sp_mod2_rat_log_nug)
```

* Did including the spatial error term have a large impact on the coefficients
of the model?

  Yes, including the spatial error term increased the coefficients in both models.
  The coefficient more than doubled for both spatial structures used in the first 
  model, though there was not a difference between results of models including
  different spatial structures. 
  In the second model, there was an increase in all models with spatial error 
  terms, but the greatest increase was in the rational model with a nugget.

* Did including the spatial error terms significantly improve model fit (use 
function `anova` to carry out model comparison)?

  Model 1 appeared to be very spatially dependent, while model 2 did not. Using
  spherical or exponential spatial structuring provided significant improvement
  to model 1 compared to one with no spatial structuring, though neither
  spherical or exponential structuring was signifcantly better than the other.
  
  Model 2 did not appear to be spatially dependent based on visual inspection.
  I played around with some spatial structures, and while there was statistically
  significant improvement with spherical or exponential structuring compared
  to the model with no spatial structure, the improvement was less than that
  of Model 1, and the results of both spherical and exponential structuring were
  identical in Model 2. 
  When I mapped out the residuals in model 2, it turned out that they were not
  normally distributed, which prompted me to log transform the data. This greatly
  improved the model, though did not change my visual interpretation of spatial 
  dependence within the model. Under the log-transformed model, the original 
  variogram still showed little visual evidence of spatial dependence, though 
  there was statistical evidence that including spatial structuring in the model 
  significantly improved the model. I tried 4 different models; rational, 
  exponential, spherical, and rational with nugget. The best model was one with 
  rational spatial structure with the nugget included.


* Explain why you did or did not observe a difference in the influence of adding 
the spatial error term between the two models.

  There was a difference in the influence of adding the spatial error term between
  the two models. The most likely reason for this is that the species included
  as predictors have different biology and might each be influenced by different
  kinds of spatial structuring. For example, it is possible that no single species
  included lacks spatial dependence. However, the kind of spatial dependence that
  each one is influenced by might be different. Species that are predators are
  more likely to be spatially structured by internal factors such as competition, 
  while other species may be structured strongly by environmental factors. Thus,
  some species could be influenced strongly by induced spatial dependence, while
  others are autocorrelated. Therefore, our model examining one species as a 
  predictor variable may be more strongly influenced by spatial structuring because
  it is more likely that only one kind spatial influence is present, which
  makes it easier to detect, and therefore easier to correct against. In the model
  with many species, different kinds of spatial influences may be causing an effect,
  and on top of that, each species included may have a different influence
  on the likelihood of our response species being present.
  
  
  
  